Large Language Models in Reinforcement Learning - A Comparison of Methods

### 2023-11-03 doc/paper.tex

% LLM policy optimization gave a NUMBER\% increase compared to the PPO baseline and a NUMBER\% increase compared to reward shaping.

\section{Conclusion}

As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve model performance. In the Minigrid environments we have tested on, we have shown that LLM reward shaping gives increased performance across the board, which shows it to be a good candidate for reward shaping in many environments with sparse rewards. Our method of policy integration has not shown similar increases, giving lower rewards on average than the baseline.

For the DoorKey environment, we found that LLM reward shaping gives a very significant $24\%$ increase in average rewards, aiding the agent in all stages of training. LLM policy influencing has sadly not given the same increase, underperforming in the more difficult environments.

For the UnlockPickup environment, we find that all methods struggle to complete the goal, but that LLM-aided methods also here outperform vanilla PPO, achieving METRIC.

>>>>

\section{Discussion, Future Work and Conclusion}

As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance. There are various ways to do this, and our work attempts to compare two main methodologies of integrating the LLM, namely through reward shaping and policy influencing. We have implemented algorithms to perform this on top of a PPO baseline, and have measured varying results across environments of different complexities.

Our reward shaping algorithm did best in the more complex environments (DoorKey and Unlock Pickup), while policy influencing performed better in the simpler Empty environment. A possible explanation for this could lie in the details of how the algorithms impact choice of action. LLM reward shaping does not directly impact decision making at all; it only impacts the reward function, meaning the mechanism that dictates which action is taken is still solely the actor network present in PPO. This means that reward shaping does not introduce any risk of the agent choosing a suboptimal action directly due to being told to do so by the LLM. Learning which action to take is still done through the same process of trial and error as in classical RL.

However, the causality behind which action is taken is different in LLM policy influencing; the LLM suggestions are added directly to the logits used to calculate probabilities of each action, meaning that the LLM directly influences the decision making process. Every time the LLM suggests performing a suboptimal action, this impacts the action probabilities, regardless whether the PPO baseline has learned that another action is better or not. Over time, this may lead to the PPO not being able to explore other actions than the ones proposed by the LLM, thus making the PPO learn slower than if it was allowed to choose actions on its own. In other words, the LLM may have too much impact in the decision making when performing policy influencing, making the algorithm prone to errors from the LLM.

A natural suggestion for how to improve the performance of LLM policy influencing in the future thus becomes to decrease the influence from the LLM over time, letting the PPO choose on its own as learning progresses. One way of doing this is through annealing of the LLM values added to the logits. This can be combined with experimenting with a hyperparameter determining how the LLM values should be weighted at the beginning of training. We have attempted to do this, but have encountered programmatic issues regarding scaling; the logits change in magnitude over time, and a fitting way of dynamically changing the LLM contributions to the logits has to be found.
